---
title: "clc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(clc)
```

## CLC

The goal of `clc` is to support operations with CLC data, addressing the challenges and specificities it presents, particularly those related to its associated styles. Its main characteristics are the following:

1. **Standardized Handling of CLC Styles**  
   CLC data includes predefined categories and associated styles, which must be correctly interpreted and applied for consistent visualization. The `clc` package automates the extraction, application, and preservation of styles across workflows.

2. **Streamlined Vector and Raster Workflows**  
   Tasks such as reading CLC vector data, applying styles, clipping layers, and converting vector data to raster format involve multiple steps. The `clc` package provides simplified, high-level functions to handle these tasks efficiently.

3. **Support for Multiple Data Sources and Targets**  
   The `clc` package handles CLC data in both GeoPackage and PostGIS formats, supporting them as sources and destinations.
